### 2017-05-08 paper/futures.tex
adaption of the paper to the new Multicore instance and the MVar Future.

instance (Trans a) => Future RemoteData a where
    put = arr (\a -> RD { rd = release a })
    get = arr rd >>> arr fetch
\end{lstlisting}
For ParMonad and Multicore we can use a basic dummy wrapper because we have shared memory in a single node:
\begin{lstlisting}[frame=htrbl]
data BasicFuture a = BF { val :: a }

instance (NFData a) => Future BasicFuture a where
    put = arr (\a -> BF { val = a })
    get = arr val
\end{lstlisting}
To fit the \code{ArrowParallel} instances we gave earlier, we also have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
We need this implementation for our \code{RemoteData} wrapper
\begin{lstlisting}[frame=htrbl]
instance NFData (RemoteData a) where
    rnf = rnf . rd
instance Trans (RemoteData a)
\end{lstlisting}
and the following for the \code{BasicFuture} dummy type:
\begin{lstlisting}[frame=htrbl]
instance (NFData a) => NFData (BasicFuture a) where
	rnf = rnf . val
\end{lstlisting}
Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
\begin{lstlisting}[frame=htrbl]
someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
someCombinator fs1 fs2 =
	parEvalN () (map (>>> put) fs1) >>>
	rightRotate >>>
	parEvalN () (map (get >>>) fs2)
\end{lstlisting}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.~3 is not really clear. Do Figs 2-3 with a lot of load?}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withFutures}
	\caption[with Futures]{Communication between 4 threads with Futures}
	\label{fig:withFutures}
\end{figure}

>>>>

instance (Trans a) => Future RemoteData a where
    put = arr (\a -> RD { rd = release a })
    get = arr rd >>> arr fetch
\end{lstlisting}
For ParMonad and Multicore we can simply use \code{MVar}s because we have shared memory in a single node:
\begin{lstlisting}[frame=htrbl]
{-# NOINLINE putUnsafe #-}
putUnsafe :: a -> MVar a
putUnsafe a = unsafePerformIO $ do
    mVar <- newEmptyMVar
    putMVar mVar a
    return mVar

instance (NFData a) => Future MVar a where
    put = arr putUnsafe
    get = arr takeMVar >>> arr unsafePerformIO
\end{lstlisting}
To fit the \code{ArrowParallel} instances we gave earlier, we also have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
We need this implementation for our \code{RemoteData} wrapper,
\begin{lstlisting}[frame=htrbl]
instance NFData (RemoteData a) where
    rnf = rnf . rd
instance Trans (RemoteData a)
\end{lstlisting}
while \code{MVar} already has a suitable NFData implementation.
\\\\
Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
\begin{lstlisting}[frame=htrbl]
someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
someCombinator fs1 fs2 =
	parEvalN () (map (>>> put) fs1) >>>
	rightRotate >>>
	parEvalN () (map (get >>>) fs2)
\end{lstlisting}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.~3 is not really clear. Do Figs 2-3 with a lot of load?}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withFutures}
	\caption[with Futures]{Communication between 4 threads with Futures}
	\label{fig:withFutures}
\end{figure}

### 2017-05-08 paper/mapSkeletons.tex
adaption of the paper to the new Multicore instance and the MVar Future.

\section{Map-based Skeletons}
\label{sec:map-skeletons}
Now we have developed Parallel Arrows further enough to define some algorithmic skeletons useful to an application programmer.
\subsection{Parallel map}
\begin{center}
	\includegraphics[scale=0.7]{images/parMap}
\end{center}
\code{parMap} is probably the most common skeleton for parallel programs. We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}. 
Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f = parEvalN conf (repeat f)
\end{lstlisting}

>>>>

\section{Map-based Skeletons}
\label{sec:map-skeletons}
Now we have developed Parallel Arrows far enough to define some algorithmic skeletons useful to an application programmer.
\subsection{Parallel map}
\begin{center}
	\includegraphics[scale=0.7]{images/parMap}
\end{center}
\code{parMap} is probably the most common skeleton for parallel programs. We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}. 
Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f = parEvalN conf (repeat f)
\end{lstlisting}

### 2017-05-08 paper/parallelHaskells.tex
adaption of the paper to the new Multicore instance and the MVar Future.

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009} is way to do parallel processing found in standard GHC.\footnote{See \url{https://hackage.haskell.org/package/parallel-3.2.1.0}.} It ships with parallel evaluation strategies for several types which can be applied with \code{using :: a -> Strategy a -> a}. For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}. This lazy list \code{[b]} is then forcibly evaluated in parallel with the strategy \code{Strategy [b]} by the \code{using} operator. This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a}.

\begin{lstlisting}[frame=htrbl]
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = zipWith ($) fs as `using` parList rdeepseq
\end{lstlisting}
\begin{center}
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
\end{center} %$ %% formatting

@@

\paragraph{Eden TraceViewer}
To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, in parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See ..... on hackage for the last available version of Eden TraceViewer. There was an effort to implement the TraceViewer using modern web technologies \cite{traceviewer-web}.} \cite{Berthold2007a}. In the next sections we will present some \emph{traces}, the post-mortem process diagrams of Eden processes and their activity.

>>>>

\subsubsection{Multicore Haskell}
Multicore Haskell \cite{Marlow2009} is way to do parallel processing found in standard GHC.\footnote{See \url{https://hackage.haskell.org/package/parallel-3.2.1.0}.} It ships with parallel evaluation strategies for several types which can be applied with \code{using :: a -> Strategy a -> a}. For \code{parEvalN} this means that we can just apply the list of functions \code{[a -> b]} to the list of inputs \code{[a]} by zipping them with the application operator \code{\$}. This lazy list \code{[b]} is then according to a \code{Strategy [b]} by the \code{using :: a -> Strategy a -> a} operator. This strategy can be constructed with \code{parList :: Strategy a -> Strategy [a]} and \code{rdeepseq :: NFData a => Strategy a} where the latter is a strategy which evalutes to normal form.
\textbf{\textcolor{red}{explain usage of pseq}}

\begin{lstlisting}[frame=htrbl]
parEvalN :: (NFData b) => [a -> b] -> [a] -> [b]
parEvalN fs as = let bs = zipWith ($) fs as in
	(bs `using` parList rdeepseq) `pseq` bs
\end{lstlisting}
\begin{center}
	\includegraphics[scale=0.5]{images/parEvalNMulticore}
\end{center} %$ %% formatting

@@

\paragraph{Eden TraceViewer}
To comprehend the efficiency and the lack thereof in a parallel program, an inspection of its execution is extremely helpful. While some large-scale solutions exist \cite{Geimer2010}, the parallel Haskell community mainly utilises the tools Threadscope \cite{Wheeler2009} and Eden TraceViewer\footnote{See ..... on hackage for the last available version of Eden TraceViewer. There was an effort to implement the TraceViewer using modern web technologies \cite{traceviewer-web}.} \cite{Berthold2007a}. In the next sections we will present some \emph{traces}, the post-mortem process diagrams of Eden processes and their activity.

### 2017-05-08 paper/parrows.tex
adaption of the paper to the new Multicore instance and the MVar Future.

\section{Parallel Arrows}
\label{sec:parallel-arrows}
We have seen what Arrows are and how they can be used as a general interface to computation. In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textbf{parallel computation} as well. We start by introducing the interface and explaining the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\begin{center}
	\includegraphics[scale=0.7]{images/parEvalN}
\end{center}
\subsection{The ArrowParallel typeclass}
As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel:
\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function \lstinline{listApp}, but does parallel instead of serial evaluation).
\begin{lstlisting}[frame=htrbl]
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{lstlisting}
With this definition of \code{parEvalN}, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{lstlisting}
Sometimes parallel Haskells require additional configuration parameters, \eg an information about the execution environment. For this reason we also introduce an additional \code{conf} parameter to the function. We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}. So we add it to the type signature of the typeclass as well.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{lstlisting}
Note that we don't require the \code{conf} parameter in every implementation. If it is not needed, we usually want to allow the \code{conf} type parameter to be of any type and don't even evaluate it by blanking it in the type signature of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.

\subsection{Multicore Haskell}
The Multicore Haskell implementation of this class is straightforward using listApp from chapter \ref{utilfns} combined with the \code{using} operator from Multicore Haskell.
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b conf where
		parEvalN _ fs = listApp fs >>> arr (flip using $ parList rdeepseq)
\end{lstlisting}% $ %% formatting
We hardcode the \code{parList rdeepseq} strategy here, as in this context it is the only one making sense, since we usually want the output list to be fully evaluated to its normal form.

\subsection{ParMonad}
The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a => a -> Par (IVar a)} function. The latter forks away the computation of a value and returns an \code{IVar} containing the result in the \code{Par} monad.

>>>>

\section{Parallel Arrows}
\label{sec:parallel-arrows}
We have seen what Arrows are and how they can be used as a general interface to computation. In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textbf{parallel computation} as well. We start by introducing the interface and explaining the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\begin{center}
	\includegraphics[scale=0.7]{images/parEvalN}
\end{center}
\subsection{The ArrowParallel typeclass}
As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel:
\begin{lstlisting}[frame=htrbl]
parEvalN :: [a -> b] -> [a] -> [b]
\end{lstlisting}
Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (This combinator is similar to our utility function \lstinline{listApp}, but does parallel instead of serial evaluation).
\begin{lstlisting}[frame=htrbl]
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{lstlisting}
With this definition of \code{parEvalN}, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we have to introduce the new typeclass \code{ArrowParallel} to host this combinator.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{lstlisting}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs. normalform). For this reason we also introduce an additional \code{conf} parameter to the function. We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}. So we add it to the type signature of the typeclass as well.
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{lstlisting}
Note that we don't require the \code{conf} parameter in every implementation. If it is not needed, we usually just default the \code{conf} type parameter to () and even blank it out in the parameter list of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.

\subsection{Multicore Haskell}
The Multicore Haskell implementation of this class is implemented in a straightforward manner by using listApp from chapter \ref{utilfns} combined with the \code{withStrategy :: Strategy a -> a -> a} and \code{pseq :: a -> b -> b} combinators from Multicore Haskell, where \code{withStrategy} is the same as \code{using :: a -> Strategy a -> a} but with flipped parameters.
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) =>
	ArrowParallel arr a b () where
    	parEvalN _ fs =
       		listApp fs >>>
        	arr (withStrategy (parList rdeepseq)) &&& arr id >>>
        	arr (uncurry pseq)
\end{lstlisting}% $ %% formatting
For most cases this fully evaluating version would probably suffice, but as the Multicore Haskell interface allows the user to specify the level of evaluation to be done via the \code{Strategy} interface, we want to the user not to lose this ability because of using our API. We therefore introduce the \code{Conf a} data-type that simply wraps a \code{Strategy a}. We can't directly use the \code{Strategy a} type here as GHC (at least in the versions used for development in this paper) does not allow type synonyms in type class instances.
\begin{lstlisting}[frame=htrbl]
data Conf a = Conf (Strategy a)
\end{lstlisting}
We simply unwrap the strategy out of this wrapper and get the following (configurable) \code{ArrowParallel} instance:
\begin{lstlisting}[frame=htrbl]
instance (NFData b, ArrowApply arr, ArrowChoice arr) => ArrowParallel arr a b (Conf b) where
    parEvalN (Conf strat) fs =
        listApp fs >>>
        arr (withStrategy (parList strat)) &&& arr id >>>
        arr (uncurry pseq)
\end{lstlisting}
\subsection{ParMonad}
The ParMonad implementation makes use of Haskells laziness and ParMonad's \code{spawnP :: NFData a => a -> Par (IVar a)} function. The latter forks away the computation of a value and returns an \code{IVar} containing the result in the \code{Par} monad.
